---
title: "Local False Discovery Rate"
author: "Zachary McCaw"
date: "Updated: 2026-03-06"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LFDR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, cache = FALSE, results = "hold")
eval_plot <- require(ggplot2) & require(reshape2)
```

# Contents

* [Background](#background)
* [Mixture Density](#mixture-density)
* [Null Density](#null-density)
* [Local FDR](#local-fdr)
* [End-to-end pipeline](#end-to-end-pipeline)
* [Summary table](#summary-table)

# Background

In large-scale testing (e.g., many genes, voxels, or hypotheses), we observe a mix of null and non-null cases. Each statistic (often converted to a z-score) is thought of as drawn from either a null distribution or an alternative distribution. The **local false discovery rate** (lfdr) at a value $\zeta$ is the posterior probability that an observation with value $\zeta$ came from the null component. So the lfdr is "local" because it depends on the exact observed value, not just on whether the statistic exceeds some threshold.

This contrasts with the **tail false discovery rate** (tFDR), which answers: among all observations as or more extreme than $\zeta$, what fraction are expected to be null? The tFDR is the expected proportion of false discoveries in the rejection region $\{Z \geq \zeta\}$ (or the analogous left tail), and is the quantity controlled by procedures like Benjamini–Hochberg when working with p-values. The local FDR is strictly about the probability that a single observation at $\zeta$ is null; the tail FDR averages over a region and is useful for choosing cutoffs and reporting error rates for sets of discoveries.

To estimate either quantity we need: (1) the overall mixture density of the observations, (2) the null distribution (location, scale, and possibly proportion), and (3) the alternative only implicitly (via the mixture and null). The rest of this vignette walks through estimating the mixture density, fitting the empirical null, and then computing the local and tail FDR curves.

# Mixture Density

## Density Estimation

Consider the following two-component normal mixture:

$$
\pi_{0}N(\mu_{0},\sigma_{0}^2) + (1-\pi_{0})N(\mu_{1},\sigma_{1}^{2})
$$

Below, $2\times 10^{3}$ observations are drawn from the two-component normal mixture with null proportion $\pi_{0}=0.95$, null parameters $(\mu_{0}=0,\sigma_{0}^{2}=1)$, and alternative parameters $(\mu_{1}=2,\sigma_{1}^{2}=1)$:

```{r}
library(LFDR)
set.seed(100)
# Data generation.
data <- RNM(n_obs = 2e3, n_comp = 2, pi = c(0.95, 0.05), loc_params = c(0, 2), scale_vars = c(1, 1))

# True component.
comp <- data[, 1]

# Observation.
z <- data[, 2]
```

An exponential family density of order $d=5$ is fit to the sample by using 1. a natural spline basis, with knots on the sample quantiles, and 2. an orthogonal polynomial basis.  

```{r}
# Fit density via natural splines.
fit_ns <- FitEF(z = z, n_bins = 100, degree = 5, method = "ns")

# Fit density via orthogonal polynomials.
fit_poly <- FitEF(z = z, n_bins = 100, degree = 5, method = "poly")
```

```{r, echo=FALSE, eval=eval_plot, fig.align="center", fig.width=5}
# Plotting fitted mixture density

# Observed data.
hist_data <- data.frame("z" = z)

# Evaluation grid.
x <- seq(from = 1.1 * min(z), to = 1.1 * max(z), length.out = 1001)
curve_df <- data.frame("x" = x, "NS" = fit_ns$f(x), "Poly" = fit_poly$f(x))
curve_df <- melt(curve_df, id.vars = "x")
colnames(curve_df) <- c("x", "Method", "y")

# Plotting.
palette <- c(rgb(230, 190, 0, max = 255), rgb(0, 120, 200, max = 255))
q <- ggplot() + theme_bw()
q <- q + geom_histogram(data = hist_data, aes(x = z, y = ..density..), fill = "gray", alpha = 0.8, bins = 50)
q <- q + geom_line(data = curve_df, aes(x = x, y = y, color = Method))
q <- q + scale_color_manual(values = palette)
q <- q + theme(legend.position = c(0.8, 0.8))
q <- q + labs("x" = "Observation", "y" = "Density")
q <- q + ggtitle("Fitted Mixture Densities")
print(q)
```

## Density Sampling

A sample $(z_{i}^{*})$ of size $n=10^{4}$ is drawn from fitted exponential family using the accept-reject (AR) method. The proposal distribution is normal, centered on the mean of the $(z_{i})$, with twice the standard deviation of the $(z_{i})$.

```{r, eval=FALSE}
# Sample fitted density.
ar_sample <- SampleEF(z = z, f = fit_poly$f, prop_mean = mean(z), prop_var = 2 * var(z), n_samp = 1e4, parallel = FALSE)
```

The AR sample $(z_{i}^{*})$ allows for estimation of moments for the fitted exponential family. To match the generative model, the AR sample should have mean $2\pi_{0} = 0.1$ and variance $1+4\pi_{0}(1-\pi_{0}) \approx 1.2$:

```{r, eval=FALSE}
# Mean of AR sample.
round(mean(ar_sample), digits = 2)

# Variance of AR sample.
round(var(ar_sample), digits = 2)
```

# Empirical Null Distribution

## Estimation

The empirical null distribution is taken as normal, however the location and scale are not required to have the theoretical values of $\mu_{0}=0$ and $\sigma_{0}^{2}=1$. Rather, these parameters, and the proportion of observations $\pi_{0}$ arising from the null density, are estimated by central matching `CM` or maximum likelihood `ML`. 

Estimation requires specification of a *null neighborhood* $A_{0} = [L,U]$. Observations falling in the null neighborhood are assumed to have arisen from the null density. In central matching, $A_{0}$ is partitioned as $L = \xi_{0} < \cdots < \xi_{K} = U$. The log mixture density $\ln\hat{f}$ is evaluated over the partition points, and the evaluations $\ln\hat{f}_{i}$ are regressed on a quadratic function of the input. In maximum likelihood, $(\mu_{0},\sigma_{0}^{2})$ are estimated by fitting a truncated normal distribution to $A_{0}$. The null proportion is obtained from $(\mu_{0},\sigma_{0}^{2})$ and the proportion of $z$ scores falling in $A_{0}$. 

```{r}
# Estimation by central matching.
null_cm <- ENull(z = z, lower_bound = -2, upper_bound = 2, method = "CM", f = fit_poly$f)

# Estimation by maximum likelihood.
null_ml <- ENull(z = z, lower_bound = -2, upper_bound = 2, method = "ML")

# Tabulate estimates.
null_tab <- round(rbind(as.numeric(null_cm), as.numeric(null_ml)), digits = 2)
colnames(null_tab) <- c("Mean", "Var", "pi0")
print(null_tab)
```

# Local FDR

Using the fitted mixture and empirical null from the previous sections, we can estimate the local and tail FDRs. The local false discovery rate (lfdr) at $\zeta$ is the posterior probability that an observation with value $\zeta$ arose from the null component of the mixture: 

$$
\text{lfdr}(\zeta) = \frac{\pi_{0}f_{0}(\zeta)}{\pi_{0}f_{0}(\zeta) + (1-\pi_{0})f_{1}(\zeta)}
$$

The tail false discovery rate (tFDR) at $\zeta$ estimates the posterior probability that an observation with value as or more extreme than $\zeta$ arose from the null component of the mixture:

$$
\text{tFDR}(\zeta) = E[\text{fdr}(Z)|Z\geq \zeta] = \frac{\pi_{0}S_{0}(\zeta)}{\pi_{0}S_{0}(\zeta)+(1-\pi_{0})S_{1}(\zeta)}
$$

The function `FDR` estimates the local and tail false discovery rate functions using the observations `z`, a fitted mixture density $f(z)=\pi_{0}f_{0}(z)+(1-\pi_{0})f_{1}(z)$, and the parameters $(\mu_{0},\sigma_{0}^{2},\pi_{0})$ of the empirical null. Two functions are returned, `lfdr` and `tFDR` for calculating the local and tail false discovery rates, respectively. 

```{r}
# Estimate FDR curves.
fdr_out <- FDR(z = z, f = fit_poly$f, n_bins = 80, null_mean = null_ml$m0, null_var = null_ml$v0, null_prop = null_ml$p0)

# Local FDR.
lfdr <- fdr_out$lfdr

# Tail FDR.
tfdr <- fdr_out$tFDR

# Evaluations for the observed data.
lfdr_vals <- lfdr(z)
tfdr_vals <- tfdr(z)
```

```{r, echo=FALSE, eval=eval_plot, fig.align="center", fig.width=5}
# Evaluation grid.
x <- seq(from = 1.1 * min(z), to = 1.1 * max(z), length.out = 1001)

# FDRs.
fdr_curve <- data.frame("x" = x, "lfdr" = lfdr(x), "tFDR" = tfdr(x))
fdr_curve <- melt(fdr_curve, id.vars = "x")
colnames(fdr_curve) <- c("x", "FDR", "y")

# Plotting.
palette <- c(rgb(230, 190, 0, max = 255), rgb(0, 120, 200, max = 255))
q <- ggplot() + theme_bw()
q <- q + geom_line(data = fdr_curve, aes(x = x, y = y, color = FDR))
q <- q + geom_hline(yintercept = 0.2, color = "gray", linetype = "dashed")
q <- q + scale_color_manual(values = palette)
q <- q + theme(legend.position = c(0.9, 0.8))
q <- q + labs("x" = "Observation", "y" = "Rate")
q <- q + ggtitle("Estimated Local and Tail FDRs")
print(q)
```

# End-to-end pipeline

The function `LFDRPipeline` runs the full workflow in one call: it fits the mixture density, estimates the empirical null, and computes the local and tail FDR. All intermediate results (`fit`, `null`, `fdr`) are returned so you can inspect or reuse them.

```{r}
# One-call pipeline (same data as above).
pipe <- LFDRPipeline(
  z = z,
  n_bins = 80,
  degree = 5,
  fit_method = "poly",
  lower_bound = -2,
  upper_bound = 2,
  null_method = "ML"
)

# Access components.
names(pipe)
pipe$null
```

The result matches the step-by-step approach: `pipe$fit` is the output of `FitEF`, `pipe$null` the output of `ENull`, and `pipe$fdr` the output of `FDR`.

# Summary table

`FDRSummary` builds a data frame that pairs each z-score with its local and tail FDR. Optionally, you can supply an lfdr cutoff to add a `discovery` flag (`TRUE` when `lfdr <= cutoff`).

```{r}
# Summary without cutoff.
summary_tbl <- FDRSummary(z = z, fdr = pipe$fdr)
head(summary_tbl, n = 5)

# Summary with discovery flag at lfdr <= 0.2.
summary_cutoff <- FDRSummary(z = z, fdr = pipe$fdr, cutoff = 0.2)
table(summary_cutoff$discovery)
head(summary_cutoff, n = 5)
```

Use the summary table to merge FDR outputs with your original data or to report the number of discoveries at a chosen cutoff.